---
title: "Worst Graph: Creative Graph-Making Exploration"
description: |
  A graph demonstrating what NOT to do, created in ESM 206 at UCSB.
author:
  - name: Erika Egg
    url: {}
date: 2022-02-21
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(extrafont) 
font_import() #may potentially be needed to use wingdings later on, if you've never used the extrafont package before
library(png)
library(grid)
```


```{r, include=FALSE, echo = TRUE, message = FALSE, warning = FALSE}
space_launches <- read_csv(here("data", "space_launches.csv")) #read in space launch data using here()
```

## Worst graph exploration
### Goal: 
Make the worst possible graph (unclear, irresponsible, terrible looking) I possibly can by customizing (from the ggplot default) in at least 7 ways - more encouraged - to make it as terrible as possible. 

```{r, echo = TRUE, message = FALSE, warning = FALSE}
launches_pegasus <- space_launches %>% 
  select(launch_year, type) %>% #select only columns launch_year and type
  filter(str_detect(type, "Pegasus")) #filter for only types containing Pegasus

max_cat <- readPNG(here("data", "max_cat.png")) #read in image for use as background

ggplot(data = launches_pegasus, aes(x = launch_year)) + 
  annotation_custom(rasterGrob(max_cat, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf) +
  geom_histogram(color = "yellow",
           aes(fill = type)) + #1 added legend where you have to look back and forth and match things up to graph; two colors are almost the same/hard to distinguish
  coord_flip() + #2 flipped it horizontally which makes it harder to understand
  scale_fill_manual(values = c("magenta1","yellow","chartreuse1","cyan1", "lawngreen")) + #3 chose very bright contrasting colors that hurt eyes
  labs(x = "# of pegasus type space launch ship shuttles going into outer space, blasting off, reaching the stars", 
       y = "the launch years, every year there was a pegasus type launch including years from 1990 to 2016 maybe idk i think that's it", 
       title = "hdfoighsdoigshertoihewoirhtoi") + #4 made axes long and with extra fluff; title is gibberish
  theme(plot.title = element_text(size = 6, family = "Wingdings"), axis.text.y = element_text(angle = 120), axis.text.x = element_text(angle = -120), axis.title.y = element_text(angle = -135, size = 5, color = "cyan1"), axis.title.x = element_text(angle = 20, size = 5, color = "magenta1")) #5 very small size font for title/axes, #6 changed title font to unreadable wingdings, #7 angled/upsidedown years on y-axis and counts on x-axis; angled x and y axis labels as well, #8 added image background of my cat
```

## Conclusion
By looking into what makes a graph bad, I learned a lot about what to include to make a graph good, such as color considerations, clarity of writing, orientation of text, and balancing simplicity with complexity.