---
title: "Visualization of California oil spill incidents in 2008"
description: |
  Spatial analysis from ESM 244 at UCSB.
author:
  - name: Erika Egg
    url: {}
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(sf)
library(here)
library(tmap)
library(janitor)
```

## Overview
The data used in this report includes locations of oil spills in California in 2008 as well as California county boundaries, which are merged together. Here I first look at where all of the 2008 oil spills were in California in an exploratory interactive visualization. Then, I construct a static choropleth map that shows which counties had the most oil spills in 2008.

### Citations:
Oil spill data from CA DFW Oil Spill Incident Tracking, 2008. https://gis.data.ca.gov/datasets/7464e3d6f4924b50ad06e5a553d71086_0/data
Shapefile data for the borders of California counties from the U.S. Census Bureau. https://data.ca.gov/dataset/ca-geographic-boundaries

## Initial wrangling
```{r}
# Read in the data
# County shapes for CA
ca_counties_sf <- read_sf(here("data", "CA_Counties_TIGER2016.shp")) %>%
   clean_names()

# Oil spill events
oil_sf <- read_sf(here("data", "ds394.shp")) %>%
   clean_names()
```

```{r}
# Select only county names (for combining with oil spill dataset later)
ca_subset_sf <- ca_counties_sf %>% 
  select(county_name = name)

# Check counties CRS
#ca_subset_sf %>% st_crs()

# Change the oil spill CRS to match counties CRS
oil_3857_sf <- st_transform(oil_sf, st_crs(ca_counties_sf))

# Then check oil spill CRS to make sure it now matches
#oil_3857_sf %>% st_crs()
```

## Exploratory interactive map: location of oil spill events
```{r, fig.align='center'}
# Set the viewing mode to "interactive"
tmap_mode(mode = "view")

# Make a map with county outlines, then add another shape layer for the oil spill records (added as dots)
tm_shape(ca_subset_sf) +
   tm_borders(lwd = 1, 
              col = "grey") +
   tm_shape(oil_3857_sf) +
   tm_dots(col="red2")
```
**Figure 1.** Exploratory map allowing users to explore the locations of individual oil spills in California. Outlines of different counties included to make locations of the spills more clear.

## Choropleth map: inland oil spills by county
```{r, fig.align='center'}
# Join the two datasets
ca_oil_sf <- ca_subset_sf %>% 
   st_join(oil_3857_sf)

# Preparing oil spill counts by county
oil_counts_sf <- ca_oil_sf %>% 
   filter(inlandmari == "Inland") %>%
   group_by(county_name) %>%
   summarize(oil_records = sum(!is.na(oesnumber)))

# Adding a 0 count row for county that had none
modoc_geom <- ca_oil_sf %>%
   filter(county_name == "Modoc") %>%
   select(geometry) %>%
   distinct()

county <- "Modoc"
count <- 0
geom <- st_transform(modoc_geom, st_crs(oil_counts_sf))
new_row <- c(county, count, geom)

oil_counts_sf_fixed <- oil_counts_sf
oil_counts_sf_fixed[nrow(oil_counts_sf) + 1, ] <- new_row

# Make map of oil spill counts by county
ggplot(data = oil_counts_sf_fixed) +
   geom_sf(aes(fill = oil_records), 
           color = "white", 
           size = 0.1) +
   coord_sf(datum = NA) +
   scale_fill_gradientn(colors = c("lightgray","orange2","red2")) +
   theme_minimal() +
   labs(fill = "Number of oil spill records",
        title = "Inland oil spills by county in California, 2008") +
   theme(text = element_text(family = "Courier",
                             size = 12,
                             face = "bold"),
         axis.text = element_text(size = 10))
```

**Figure 2.** Choropleth map showing the number of inland oil spill events in each county. The more red a county is, the higher the number of oil spills. Los Angeles county contains the most inland spills.


