---
title: "Bonanza Hares: Exploratory Report"
description: |
  A sample report constructed in ESM 206 at UCSB.
author:
  - name: Erika Egg
    url: {}
date: 2022-02-17
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(ggbeeswarm)
library(kableExtra)
library(effsize)
```

## Introduction

This report provides an exploratory overview of snowshoe hare data from near Bonanza Creek Experimental Forest at three sites (Black Spruce, Mature, and Riparian) (Kielland et al. 2017). The Black Spruce site refers to a mature lowland black spruce stand, while the Mature and Riparian sites includes stands of a variety of mature tree species, shrub species, and/or grass species (differing by site, but including: white spruce, black spruce, paper birch, willow, prickly rose, fireweed, among others), with a key difference being that the Riparian is located in a flood zone (Kielland et al. 2017). The data used here spans 1999 to 2012 (Kielland et al. 2017). Specifically, we conduct analyses of juvenile snowshoe hare trappings over the timespan (including only trappings from the three sites described) and of juvenile snowshoe hare weights in relation to sex, site, and hind foot length.

## Data and analyses

Snowshoe hare measurements (such as weight and hind foot length, key focuses in this report) are provided by the Bonanza Creek LTER, a partnership between the University of Alaska Fairbanks, and the U.S. Forest Service (Kielland et al. 2017). The data contain observations for 3,197 recorded trappings, collected from 1999 to 2012 (Kielland et al. 2017). Here we utilize R and R Studio to analyze the data through the generation of ggplot visualizations (ie. beeswarm, column chart, scatterplot), descriptive statistics (ie. mean, median, standard deviation), and statistical tests and related statistics (ie. two-sample t-test with a significance level ($\alpha$) of 0.05, Cohen's *d*, Pearson's *r*, simple linear regression).

## Exploratory findings 

### A. Annual juvenile hare trap counts.

```{r, include = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
hares <- read_csv(here("data", "bonanza_hares.csv")) %>% #reading in the data
  clean_names() #cleaning the column names
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#getting the counts for the total number of juvenile (age == "j") hare trappings during each year of the study (include all grid sites)
juvenile_hares <- hares %>%
  filter(age == "j") %>% #filtering for juvenile
  mutate(date_new = mdy(date)) %>% #convert date to workable (date) format in new column
  mutate(year = year(date_new))  #extract year from new date column

count_jh <- juvenile_hares %>% #use the subset created above
  group_by(year) %>% #group by year
  count() #take the count

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
#create a finalized data visualization of the counts by year
ggplot(data = count_jh, aes(x = year, y = n)) + #setting x axis as year and y as the count
  geom_col(color = "black",
           fill = "slategray4") + #change fill color and edge color
  theme_bw() + #choose theme
  theme(panel.grid.minor.x=element_blank(),
           panel.grid.major.x=element_blank()) + #get rid of vertical grid lines
  labs(x = "Year",
       y = "Number of juvenile hare trappings",
       title = "Juvenile snowshoe hare trap counts by year, 1999-2012") #label axes and title

summary_count_jh <- count_jh %>%
  ungroup() %>% #ungroup the groups to get stats across all
  summarize(
    mean_trapped = mean(n), #get mean
    median_trapped = median(n), #get median
    min_trapped = min(n), #get min
    max_trapped = max(n) #get max
  )
```

**Figure 1.** Juvenile snowshoe hair trappings each year from 1999 to 2012. Height of columns indicate number of trappings in that year. Data: Kielland et al. 2017.

The minimum count of juvenile hare trappings (when excluding the two years that had 0: hares were actually trapped and recorded in 2002 and 2009 but ages weren't recorded so the data shows those years as 0 (Thanks to Joe Walderman for pointing this out). Including those 0's would then potentially skew the data inaccurately when reporting descriptive statistics, so I have excluded them) was `r summary_count_jh$min_trapped` in 2010 while the maximum was `r summary_count_jh$max_trapped` hares in 1999. Further, the mean annual number of juvenile hares trapped was `r summary_count_jh$mean_trapped` while the median was `r summary_count_jh$median_trapped` hares. Overall, as the year increases, we see a general trend of decreasing annual juvenile hare trappings. Moving forward, we suggest a comparison not only between juvenile hare trappings and year, but also factoring in effort by looking at success rate (juveniles trapped divided by the number of traps set up each year); if counts are higher in years with more traps set up in total, then it may be better to estimate changes in count by looking at how the success rates (%) change each year (ie. if 50 hares were caught in both 2011 and 2012, but in 2011 there were 1000 traps set up vs. 100 in 2012, factoring that success rate in would tell a very different story than looking at absolute counts).


### B. Visualizing juvenile hare weights by sex & site.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
#set up data to be used in visualizations
#for beeswarm
wjh <- juvenile_hares %>%
  group_by(sex, grid) %>% #group by sex and grid
  rename(Sex = sex) %>% #renaming so it looks nice on the ggplot
  mutate(Sex = case_when( #changing sex value names so they look nice on the ggplot
        Sex == 'f' ~ 'Female',
        Sex == 'm' ~ 'Male',
        Sex == 'NA' ~ 'NA'
    )) %>%
  mutate(grid = case_when( #changing grid value names so they look nice on the ggplot
        grid == 'bonbs' ~ 'Black Spruce',
        grid == 'bonmat' ~ 'Mature',
        grid == 'bonrip' ~ 'Riparian'
    ))

#for point and errorbar
summary_wjh_by_gs <- wjh %>%
  dplyr::summarize( #create summary
    mean_weight = round(mean(weight, na.rm = TRUE), 2), #of mean weight, rounded to 2 decimal places
    sd_weight = round(sd(weight, na.rm = TRUE), 2), #of standard deviation of weight, rounded to 2 decimal places
    n = n() #of sample sizes
  ) 

#summary_wjh_by_gs <- summary_wjh_by_gs[-c(7),] (if i wanted to remove the row with all NAs)

#create visualization of juvenile hare weights by sex and site (grid)  
ggplot() +
  geom_beeswarm(data = wjh, #create beeswarm plot
       aes(x = Sex, #sex on x axis
           y = weight, #weight on y axis
           color = Sex, #color varies by sex
           ),
                size = 2.5, #point sizes = 2.5
                pch = 10) + #change point shape to crosshair-like circles
  geom_point(data = summary_wjh_by_gs, aes(x = Sex, y = mean_weight), #layer with a point showing each mean weight by sex and grid
             color = "black", #set mean point color to black
             size = 1.5) + #set point size to 1.5
  geom_errorbar(data = summary_wjh_by_gs, #add layer to hraph with error bars
                aes(x = Sex, #x axis is sex like the other layers
                    ymin = mean_weight - sd_weight, #y axis represents standard deviations
                    ymax = mean_weight + sd_weight), #in both directions
                color = "black", #bars are black
                width = 0.1) + #bar width is 0.1
  theme_bw() + #change theme
  facet_wrap(~grid) + #facet wrap to separate by grid
  labs(x = "Sex",
       y = "Weight (grams)",
       title = "Juvenile hare weights by sex and site, 1999-2012") #label axes and title
  


  
```

**Figure 2.** Weight observations (g) for juvenile male and female (including NA) snowshoe hares across three sites (Black Spruce, Mature, and Riparian). Red (Female), teal (Male) and grey (NA) crosshair points indicate individual observations for weight (g) of a juvenile snowshoe hare. Errorbar endpoints indicate the standard deviation in both directions from the mean (the black point) of each sex and site pairing. Data: Kielland et al. 2017. 

By site, Black Spruce seems to have juveniles hares with higher weights (x̄ male = `r summary_wjh_by_gs$mean_weight[4]` g, x̄ female = `r summary_wjh_by_gs$mean_weight[1]` g), while the other two sites have juvenile hares with slightly lower weights (Mature: x̄ male = `r summary_wjh_by_gs$mean_weight[5]` g, x̄ female = `r summary_wjh_by_gs$mean_weight[2]` g; Riparian: x̄ male = `r summary_wjh_by_gs$mean_weight[6]` g, x̄ female = `r summary_wjh_by_gs$mean_weight[3]` g). By sex, males seem to weigh slightly more at each individual site relative to the females (x̄ male =`r summary_wjh_by_gs$mean_weight[4]` g, `r summary_wjh_by_gs$mean_weight[5]` g, `r summary_wjh_by_gs$mean_weight[6]` g vs. x̄ female = `r summary_wjh_by_gs$mean_weight[1]` g, `r summary_wjh_by_gs$mean_weight[2]` g, `r summary_wjh_by_gs$mean_weight[3]` g) but the difference doesn't appear to be significant (the exception being the NA sex at the Riparian site, but I don't believe this difference necessarily says anything meaningful, especially considering the small sample size (n = `r summary_wjh_by_gs$n[9]` hares)).


### C. Weight comparison between juvenile male & female snowshoe hares.

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) for juvenile male and female snowshoe hares. Data: Kielland et al. 2017.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#create summary table
summary_wjh_by_s <- juvenile_hares %>%
  group_by(sex) %>% #group by sex
  filter(!is.na(sex)) %>% #take out NAs for sex
  dplyr::summarize( #create summary
    mean_weight = round(mean(weight, na.rm = TRUE), 2), #of mean weight, rounded to 2 decimal places
    sd_weight = round(sd(weight, na.rm = TRUE), 2), #of standard deviation of weight, rounded to 2 decimal places
    n = n() #of sample sizes
  ) %>%
  mutate(sex = case_when( #changing sex value names so they look nice in the table
        sex == 'f' ~ 'Female',
        sex == 'm' ~ 'Male',
        sex == 'NA' ~ 'NA'
    ))

#renaming columns to make name formatting nicer
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "mean_weight"] <- "Mean Weight (g)"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "sd_weight"] <- "Standard Deviation (g)"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "n"] <- "Sample Size"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "sex"] <- "Sex"

#make the formatting of the table nicer
formatted_summary_wjh_by_s <- summary_wjh_by_s %>%
  kbl() %>%
  kable_classic_2(full_width = F) #make overall formatting nicer
formatted_summary_wjh_by_s
```
```{r, include = FALSE}
#renaming again to make coding easier
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "Mean Weight (g)"] <- "mean_weight"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "Standard Deviation (g)"] <- "sd_weight"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "Sample Size"] <- "n"
names(summary_wjh_by_s)[names(summary_wjh_by_s) == "Sex"] <- "sex"

#actual difference in means
diff_means <- summary_wjh_by_s$mean_weight[2]-summary_wjh_by_s$mean_weight[1] #find difference in mean weights from male to female

#effect size (cohen's d)
jhares_f <- juvenile_hares %>% 
  filter(sex == "f") %>% #filter for females
  filter(!is.na(weight)) %>% #take out NAs
  pull(weight) # get weight data

jhares_m <- juvenile_hares %>% 
  filter(sex == "m") %>% #filter for males
  pull(weight) #get weight data

jhares_cohen <- cohen.d(jhares_f, jhares_m) #find cohen's d
```
```{r, include=FALSE}
#graphs to explore assumptions of t-test
##histogram
ggplot(data = juvenile_hares, aes(x = weight)) + #histograms by sex looking at weight distributions
  geom_histogram(bins = 10) +
  facet_wrap(~sex) #looks fairly normal, especially for females, males looks slightly more like it could have two peaks

##QQ plot
ggplot(data = juvenile_hares, aes(sample = weight)) + #qqplots by sex looking at weight distributions
  geom_qq() +
  facet_wrap(~sex) #this is slightly curved but still looks relatively normal
```
```{r, include = FALSE}
#outcome of a two-sample t-test
jhares_ttest <- t.test(jhares_f, jhares_m) #two sample t-test
jhares_ttest_out <- broom::tidy(jhares_ttest)
jhares_ttest_out
```

The actual difference in means is `r diff_means` g, with small-moderate effect size (Cohen's *d* = `r round(jhares_cohen$estimate,4)`). Therefore, there may not be that significant of a difference in weights between juvenile male and female snowshoe hares, despite demonstrating statistical significance (t(`r round(jhares_ttest_out$parameter,2)`) = `r round(jhares_ttest_out$statistic,2)`, *p* = `r round(jhares_ttest_out$p.value,4)`).

### D. Relationship between juvenile weight & hind foot length.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
ggplot(data = juvenile_hares, 
       aes(x = hindft, y = weight)) + #select x as hindft and y as weight
  geom_point(color = "slategray4", #chose a color, color by sex doesn't reveal anything particularly interesting
                size = 1.5, #change size of points to 1.5
                pch = 10, #change point shape to crosshair-like circles
                ) +
  theme_bw() + #change theme
  labs(x = "Weight (grams)",
       y = "Hind foot length (millimeters)",
       title = "Juvenile hare weights (g) vs. hind foot length (mm), 1999-2012") #label axes and title
  
```

**Figure 3.** Relationship between weight (g) and hind foot length (mm) for all juvenile snowshoe hares. Points indicate individual snowshoe hare measurements. Data: Kielland et al. 2017.

#### Linear regression in relation to weight and hind foot length
```{r, include = FALSE}
#linear model
jhares_lm <- lm(weight ~ hindft, data = juvenile_hares)

#return the complete overview
summary_jhares_lm <- summary(jhares_lm)

#tidied return - for getting coefficient
jhares_lm_tidy <- broom::tidy(jhares_lm)

#tidied return - for getting r squared
jhares_lm_out <- broom::glance(jhares_lm)

#pearson's r
cor_jhares <- cor.test(juvenile_hares$hindft, juvenile_hares$weight)

```
The output of the linear model indicates that, on average, we expect hind foot length to increase by `r round(jhares_lm_tidy$estimate[2],4)` mm for each 1 gram increase in weight. However, the r-squared value of `r round(jhares_lm_out$r.squared,3)` indicates that only `r round(jhares_lm_out$r.squared,3)*100`% of variance in weight is explained by hind foot length, with the Pearson's *r* correlation of `r round(cor_jhares$estimate,4)` further demostrating a low-medium positive correlation. From looking at the shape of the graph, there is some concern about the linearity of the relationship and heteroscedasticity of the errors, which means that the linear model may not be the best fit to describe the relationship.

## Summary

Overall, we have noticed that:

* Annual juvenile hare trappings seem to have decreased over time in general.
* Juvenile hare weights seem to vary by site and perhaps by gender within a site (but not by gender across sites).
* Juvenile hare weight and hind foot length seem to have a positive relationship, but a linear model may not best describe this relationship.

## Citations

Kielland, Knut; Chapin, F. Stuart; Ruess, Roger W. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present, Bonanza Creek LTER - University of Alaska Fairbanks. BNZ:55, http://www.lter.uaf.edu/data/data-detail/id/55. doi:10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14


