---
title: "Wrangling and Visualizing Mountain Yellow-legged Frog (Rana muscosa) Counts"
description: |
  Creating a compound figure in ESM 244 at UCSB.
author:
  - name: Erika Egg
    url: {}
date: 2022-02-21
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readxl)
library(tidyverse)
library(lubridate)
library(patchwork)
library(extrafont) #need font_import() if first time
library(here)
```


### Introduction
The data used in this report includes amphibian abundance recorded by the Sierra Lakes Inventory Project from 1995-2002, accompanied by additional characteristics such as species, life stage, and locational attributes. Here I attempt to answer two questions: "What is the mountain yellow-legged frog abundance trend over this time period by life stage?" and "Which lakes contained the highest total observed mountain yellow-legged frog counts over this time period?" utilizing a compound figure as a final output.

### Preparatory steps
#### 1. Read in the file containing the Sierra amphibian data.
```{r}
frogs <- read_excel(here("data", "sierra_amphibians.xlsx")) #read in the data we will be using: sierra                                                                       amphibians
```


#### 2. Perform some initial wrangling on the data that can aid the creation of both plots.
```{r}
ramu_wrangled <- frogs %>%
   filter(amphibian_species == "RAMU", amphibian_life_stage != "EggMass") %>% #filter for RAMU and exclude                                                                                     EggMass stage
   mutate(date_new = ymd(survey_date)) %>% #convert date to workable (date) format in new column
   mutate(year = year(date_new)) %>% #extract year from new date column
   drop_na(amphibian_number) #remove NAs from relevant column
```

### Creating our first plot
#### 3a. Wrangle the data for the first plot by getting RAMU counts after grouping by life stage and year.
```{r}
count_ramu_stage <- ramu_wrangled %>% #use the subset created above
  group_by(amphibian_life_stage, year) %>% #group by life stage and year
  summarize(stage_frog_count = sum(amphibian_number)) #take the count of frogs
```


#### 3b. Create a graph of total mountain yellow-legged frog (Rana muscosa) count each year across all water bodies, by life stage excluding the ‘EggMass’ level.
```{r}
p1 <- ggplot(data = count_ramu_stage, aes(x = year, #set x as year,
                                          y = stage_frog_count, #y as count,
                                          color = amphibian_life_stage)) + #and color as life stage
  geom_line(size = 0.4) + #change line width
  facet_wrap(~amphibian_life_stage) + #facet wrap by life stage for readability
  scale_color_manual(values = c("palegreen3", "steelblue1", "tan")) + #change line colors
  theme_dark() + #choose theme
  theme(panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank(), #get rid of vertical grid lines 
        text=element_text(size=10,  family="Impact")) + 
  labs(x = "Year", #label x,
       y = "Yellow-legged frog counts", #and y axis,
       color = "Life Stage") #and color key
```

### Creating our second plot
#### 4a. Wrangle the data for the second plot by getting the top 5 RAMU counts (excluding Tadpoles too) after grouping by lake.
```{r}
ramu_no_tadpole <- ramu_wrangled %>%
   filter(amphibian_life_stage != "Tadpole") #filter out Tadpoles

count_ramu_lake <- ramu_no_tadpole %>%
   group_by(lake_id) %>% #group by lake id
   summarize(lake_frog_count = sum(amphibian_number)) %>% #take the counts 
   slice_max(lake_frog_count, n = 5) %>% #take only top 5 lakes, including ties if applicable
   mutate(id_reformatted = paste("Lake", lake_id)) #have a reformatted lake id
```


#### 4b. Create a column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts.
```{r}
p2 <- ggplot(data = count_ramu_lake, aes(x = fct_reorder(id_reformatted, lake_frog_count), #set x as lake id                                                                                                reordered by count
                                         y = lake_frog_count)) + #set y as count
  geom_col(color = "tan", #change edge color, 
           fill = "aquamarine4") + #and fill color 
  theme_dark() + #choose theme
  theme(panel.grid.minor.y=element_blank(),
        panel.grid.major.y=element_blank(), #get rid of horizontal grid lines 
        text=element_text(size=10,  family="Impact")) + #change text font and size
  labs(y = "Total yellow-legged frog counts (Adult + SubAdult)", #label y,
       x = "Lake ID") + #and x axis
  coord_flip() #make horizontal
```

### Final compound plot
#### 5. Put together final plot including both individual plots constructed previously.
```{r, fig.align = "center"}
p1 / p2 #create top and bottom combined plot
```
**Figure 1.** Top graph: yellow-legged frog counts each year by life stage from 1995 to 2002. Height of lines indicate number of yellow-legged frogs counted in that year for each life stage. Bottom graph: cumulative yellow-legged frogs counts across the years 1995 to 2002 by lake. The lakes with the top five counts have been included, identified only by ID numbers for confidentiality. Data: Knapp et al. 2020. 


### Citation
Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3