---
title: "Word frequency and sentiment analysis of Cyrano de Bergerac"
description: |
  Text analysis from ESM 244 at UCSB.
author:
  - name: Erika Egg
    url: {}
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(tidytext)
library(textdata)
library(pdftools)
library(ggwordcloud)
library(here)
library(jpeg)
library(grid)
library(viridis)
```

## Overview
The data in this report is sourced from a PDF (from Project Gutenberg) of the play Cyrano de Bergerac by Edmond Rostand. After removing text not part of the original play, I first explore the most frequent words throughout the spoken dialogue of the play (excluding the labels of which character is speaking). I then transition to sentiment analysis of the play by each major scene (1-5).

### Citations:
Edmond Rostand. 1897. Cyrano de Bergerac. PDF EBook version by Project Gutenberg, released 1998 and posted 2009. https://www.gutenberg.org/files/1254/1254-h/1254-h.htm

## Wrangling the text
```{r}
# Read in pdf of Cyrano de Bergerac
cyrano_text <- pdf_text(here::here('data', 'cyrano.pdf'))
```

```{r}
# Extract lines from the play
cyrano_lines <- data.frame(cyrano_text) %>%
   mutate(page = 1:n()) %>%
   mutate(text_full = str_split(cyrano_text, pattern = '\\n')) %>%
   unnest(text_full) %>%
   mutate(text_full = str_trim(text_full))

# Identify lines by scene in the play
cyrano_scenes <- cyrano_lines %>%
   slice(-(9388:9754)) %>%
   slice(-(1:157)) %>%
   mutate(scene = ifelse(str_detect(text_full, "Scene"), text_full, NA)) %>%
   fill(scene, .direction = 'down') %>%
   separate(col = scene, into = c("scene", "scene_num_full"), sep = " ") %>%
   mutate(main_scene = str_extract(scene_num_full, "[:digit:]")) %>%
   mutate(text_full = str_replace(text_full, "^.*:$", "")) # took out name mentions that are not spoken by other characters

# Get words from play
cyrano_words <- cyrano_scenes %>%
   unnest_tokens(word, text_full) %>%
   select(-cyrano_text, -scene, -scene_num_full) %>%
   drop_na(main_scene)

# Remove stop words
cyrano_words_clean <- cyrano_words %>%
   anti_join(stop_words, by = 'word')

# Get counts overall for entire play
entire_nonstop_counts <- cyrano_words_clean %>%
   count(word)

# Get the top 100 words in the play
play_top100 <- entire_nonstop_counts %>%
   filter(word != "1",
          word != "2",
          word != "3",
          word != "4",
          word != "5",
          word != "scene",
          word != "de",
          word != "bergerac") %>% # remove more non-relevant words
   arrange(-n) %>%
   slice(1:100) 
```

## Creating a word cloud
```{r, fig.align='center'}
# Read in picture for word cloud background
cyrano_pic <- jpeg::readJPEG(here("data", "cyrano_pic.jpg"))
# Link to image on web - https://www.deviantart.com/protvscar/art/Cyrano-de-Bergerac-Fly-leaf-33171255

# Make word cloud of top 100 words in the play
ggplot(data = play_top100, aes(label=word)) +
   annotation_custom(rasterGrob(cyrano_pic, 
                                width = unit(1,"npc"), 
                                height = unit(1,"npc")), 
                                -Inf, Inf, -Inf, Inf) +
   geom_text_wordcloud(aes(color = n, size = n), text=element_text(size=10,  family="Impact")) +
   scale_size_area(max_size = 7) +
   scale_color_viridis(option = "rocket", direction = -1)
```

**Figure 1.** Word cloud showing the top 100 words in the play Cyrano de Bergerac. In general, the darker the color of the word and closer to the center it is, the more times it appears in the play (most to least occurrences: purple, purple-pink, pink-red, orange, lighter shades of orange, and then white). "ah," "ay," and "love" along with key character names (Cyrano, Roxane, and Christian) are the most common. Background image created by protvscar on DeviantArt (link provided in the code above).

## Performing sentiment analysis
```{r, fig.align='center'}
# Get sentiments for words in play using afinn
cyrano_afinn <- cyrano_words_clean %>%
   inner_join(get_sentiments("afinn"), by = 'word')

# Get the counts by scene
afinn_counts <- cyrano_afinn %>%
   count(main_scene, value)

# Find the mean afinn score by scene
afinn_means <- cyrano_afinn %>%
   group_by(main_scene) %>%
   summarize(mean_afinn = mean(value))

# Create sentiment column graph by scene
ggplot(data = afinn_means, aes(x = as.numeric(main_scene), y = mean_afinn, fill = mean_afinn)) +
  geom_col() +
  theme_bw() +
  coord_flip() +
  scale_fill_gradient2(low = "tomato4", 
                       mid = "snow", 
                       high = "darkolivegreen4") +
  scale_x_continuous(trans = "reverse") +
  theme(legend.position = "none") +
  labs(x = "Scene", 
       y = "Average Afinn Sentiment Score",
       title = "Average Afinn sentiment scores by main scene") +
  theme(panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank(), #get rid of vertical grid lines
        panel.grid.minor.y=element_blank(),
        panel.grid.major.y=element_blank(),
        text=element_text(size = 10, face = "bold"))
```

**Figure 2.** Average Afinn sentiment scores by major scene in Cyrano de Bergerac. From Scene 2 onwards, the play becomes more negative. If one has read the play, one's opinion would likely be similar to the sentiments found in this chart.